---
title: "Reproducible Research, Project 1"
author: "FishTales4967"
date: "September 13, 2016"
output: html_document
---
Report on Activity Observations
===============================


## Loading and preprocessing the data


```{r hide=TRUE}
library(dplyr)
library(ggplot2)
library(lattice)
setwd("I:/rprog-Course5-Week2")
```
```{r}
tab1 <- read.csv(file="activity.csv",sep=",",stringsAsFactors = FALSE) %>% mutate(obsDate=as.Date(date,format="%m/%d/%Y"))
```


## What is mean total number of steps taken per day?


  Create Histogram of mean steps per day when missing values are excluded.

```{r}
by_date <- tab1 %>% group_by(obsDate) %>% summarize(totsteps=sum(steps))
g <- ggplot(tab1, aes(obsDate))
maintitle <- "Total Daily Steps, October - November 2012 "
mn_steps1 <- round(mean(by_date$totsteps,na.rm=TRUE),2)
md_steps1 <- round(median(by_date$totsteps,na.rm=TRUE),2)
g+ geom_histogram(aes(weight=steps),bins=61)+ geom_hline(yintercept=mn_steps1,col="blue") + xlab(" ") +ylab("Total Steps") + ggtitle(maintitle) 
```
```{r echo=FALSE}
paste("mean steps = ", mn_steps1 )
paste("median steps = ", md_steps1)
```
## What is the average daily activity pattern?
 
   Create Time Series Plot of average daily activity pattern

```{r }
by_intvl <- tab1 %>% group_by(interval) %>% summarize(int_mn=mean(steps,na.rm=TRUE))
with(by_intvl,xyplot(int_mn~interval,type="l",ylab="mean steps",main="Average Daily Step Pattern"))
row_of_max <- with(by_intvl,which.max(int_mn))
max_int <- by_intvl[row_of_max,"interval"]
step_max <- round(by_intvl[row_of_max,"int_mn"],2)
```
```{r echo=FALSE}
paste("Maximum average daily steps is ", step_max, "and occurs at interval ",max_int)
```
## impute missing values and make new histogram
To impute missing values, the average of non-missing values was computed at each interval. Then the appropriate mean value was substituted for each missing value.

```{r echo=FALSE}
paste("Count of missing step values = ", (table(is.na(tab1$steps))[2]))
```
```{r }
int_mn <- filter(tab1,steps>=0) %>% group_by(interval) %>% summarize(replc=mean(steps,na.rm=TRUE))
tab2 <- merge(tab1,int_mn)
tab3 <- tab2 %>% mutate(steps1=ifelse(is.na(steps),replc,steps))
by_date <- tab3 %>% group_by(obsDate) %>% summarize(totsteps=sum(steps1))
k <- ggplot(tab3, aes(obsDate))
maintitle <- "Total Daily Steps, Oct-Nov 2012, mean imputed for missing values "
md_steps <- round(median(by_date$totsteps,na.rm=TRUE),2)
mn_steps <- round(mean(by_date$totsteps,na.rm=TRUE),2)
k + geom_histogram(aes(weight=steps1),bins=61)+ geom_hline(yintercept=mn_steps,col="blue") + xlab(" ") + ylab("Total Steps") + ggtitle(maintitle)
```


WIth the imputed values, the mean is equal to the median.

```{r echo=FALSE}
paste("Mean steps after substituting imputed values for NAs = ", mn_steps)
paste("Median steps after substituting imputed values for NAs = ", md_steps)
```                
## Are there differences in activity patterns between weekdays and weekends?
 
 Two panel plot, weekend vs weekday
```{r }
tab4 <- tab3 %>% mutate(wkdayind=ifelse(weekdays(obsDate) %in% c("Sunday", "Saturday"),"Weekend","Weekday"))
by_intvl2 <- tab4 %>% group_by(interval,wkdayind) %>% summarize(stepsMn=mean(steps1))
ylabel <-"Mean Steps"
with(by_intvl2,xyplot(stepsMn~interval|wkdayind,type="l",ylab=ylabel,layout=(c(1,2))))
```

The pattern of activity across intervals appears to differ depending on whether the days are week days or weekend days. Week day activity appears to be higher in the 500 to 900 interval number range relative to weekends, and lower in the 1000 to 1800 interval range.
